---
title: "DATA 605: Fundamentals of Computational Mathematics"
author: "Walt Wells, Spring 2017"
subtitle: "Assignment 3"
output:
  html_document:
    css: ../custom.css
    highlight: zenburn
    theme: lumen
---
# Problem 1

### (1) What is the rank of the matrix A?

```{r}
A <- matrix(c(1,2,3,4,
              -1,0,1,3,
              0,1,-2,1,
              5,4,-2,-3), 4, byrow=T)

qr(A)$rank

## confirm
round(A %*% solve(A))
```

We have to do a little row exchange to see it, but this matrix has 4 pivots.  We also confirm that since it is invertible, the rank is the same as the dimension. 

### (2) Given an mxn matrix where m > n, what can be the maximum rank? The minimum rank, assuming that the matrix is non-zero?

If m > n then the maximum rank is n; assuming the matrix has at least one non-zero element, its minimum rank is 1.

### (3) What is the rank of matrix B?

```{r}
B <- matrix(c(1,2,1,
              3,6,3,
              2,4,2), 3, byrow=T)

qr(B)$rank
```

Since each of the rows and columns of the matrix are not linearly independant, our rank is only 1, the minimum potential rank.  

# Problem 2

Compute the eigenvalues and eigenvectors of the matrix A. You'll need to show your work. You'll need to write out the characteristic polynomial and show your solution.

### Compute Eigenvalues

1) ${det}\left(A-\lambda \,I\right)\,=\,0$

2) ${det}\left(\begin{bmatrix}1&2&3\\0&4&5\\0&0&6\end{bmatrix} - \begin{bmatrix}\lambda&0&0\\0&\lambda&0\\0&0&\lambda\end{bmatrix}\right)=0$

3) ${det}\left(\begin{bmatrix}1-\lambda &2&3\\0&4-\lambda &5\\0&0&6-\lambda \end{bmatrix}\right)=0$

4) $\left(1-\lambda\right) [\left(4-\lambda \right)\left(6-\lambda \right)-5\times 0] + 2[0 \times \left(6-\lambda \right)-5\times 0] + 3[0\times0 - 0\left(6-\lambda \right)]=0$

5) $\left(1-\lambda\right)\left(4-\lambda \right)\left(6-\lambda \right)=0$

    + _Characteristic Polynomial:_ $-\lambda^3 + 11\lambda^2 - 34\lambda + 24=0$

6) $\lambda_1=6, \lambda_2=4, \lambda_3=1$

### Compute Eigenvectors

#### $\lambda_1=6$

1) $(A- \lambda_1I)v=0$

2) $\begin{bmatrix}-5&2&3\\0&-2&5\\0&0&0\end{bmatrix}\begin{bmatrix}X\\Y\\Z\end{bmatrix}=0$

3) This becomes a system of equations such that:   

    $-5x + 2y + 3z = 0 \\ -2y + 5z = 0$

4) If we make $Y=1$:

    $-2 + 5z = 0; z = 2/5$
    
5) Substituting Y and Z:

    $-5x + 2 + 3(2/5) = 0 \\ -5x = -16/5 \\ x = 16/25$
    
6) $E_{\lambda_1=6}=c\begin{bmatrix}16/25 \\ 1 \\ 2/5\end{bmatrix}$

#### $\lambda_2=4$

1) $(A- \lambda_2 I)v=0$

2) $\begin{bmatrix}-3&2&3\\0&0&5\\0&0&2\end{bmatrix}\begin{bmatrix}X\\Y\\Z\end{bmatrix}=0$

3) This becomes a system of equations such that:   

    $-3x + 2y + z = 0\\ 5z = 0\\2z = 0$

4) If we set $Y=1$ then $X=2/3$
    
5) $E_{\lambda_2=4}=c\begin{bmatrix}2/3 \\ 1 \\ 0\end{bmatrix}$

#### $\lambda_3=1$

1) $(A- \lambda_3 I)v=0$

2) $\begin{bmatrix}0&2&3\\0&3&5\\0&0&5\end{bmatrix}\begin{bmatrix}X\\Y\\Z\end{bmatrix}=0$

3) This becomes a system of equations such that:   

    $2y + 3z = 0\\ 3y + 5z = 0\\5z = 0$

4) The only non-zero solution is $X=1$ and Y and Z = 0
    
5) $E_{\lambda_3=1}=c\begin{bmatrix}1 \\ 0 \\ 0\end{bmatrix}$


### Test

```{r}
A <- matrix(c(1,2,3,
              0,4,5,
              0,0,6), 3, byrow=T)
eigen(A)

# double check by converting to unit vectors
unit <- function(x) {
    x / sqrt(sum(x^2))
}
l1 <- unit(c(16/25, 1, 2/5))
l2 <- unit(c(2/3, 1, 0))
l3 <- unit(c(1, 0, 0))

eigenvectors <- cbind(l1, l2, l3)
eigenvectors
```