---
title: "DATA 605: Fundamentals of Computational Mathematics"
author: "Walt Wells, Spring 2017"
subtitle: "Assignment 1"
output:
  html_document:
    css: ../custom.css
    highlight: zenburn
    theme: lumen
---
# Problem 1

```{r}
u <- c(0.5, 0.5)
v <- c(3, -4)

# 1)  Calculate the dot product u.v where u = [0.5; 0.5] and v = [3;−4]
dot <- u %*% v
dot 
```

```{r}
# 2)  What are the lengths of u and v?  Please note that the mathematical notion of the length of a vector is not the same as a computer science definition.
ul <- sqrt(sum(u^2))
ul
vl <- sqrt(sum(v^2))
vl
```

```{r}
# 3) What is the linear combination:  3u − 2v?
3*u - 2*v
```

```{r}
# 4) What is the angle between u and v?
rad <- acos(dot / (ul * vl))
# in degrees
rad * 180/pi
```

# Problem 2
Set up a system of equations with 3 variables and 3 constraints and solve for x.  Please write a function in R that will take two variables (matrix A & constraint vector b) and solve using elimination.  Your function should produce the right answer for the system of equations for any 3-variable, 3-equation system.  You don’t have to worry about degenerate cases and can safely assume that the function will only be tested with a system of equations that has a solution. Please note that you do have to worry about zero pivots, though. Please note that you should not use the built-in function solve to solve this system or use matrix inverses.  The approach that you should employ is to construct an Upper Triangular Matrix and then back-substitute to get the solution.  Alternatively, you can augment the matrix A with vector b and jointly apply the Gauss Jordan elimination procedure.

Please test it with the system below and it should produce a solution
x = [−1.55, −0.32, 0.95]

## Test Matrices

```{r}
A <- matrix(c(1,1,3,2,-1,5,-1,-2,4), 3, byrow=T)
b <- matrix(c(1,2,6))
```

## Helper Functions

```{r}
# Create augmented matrix
combine <- function(A,b){
    Aug <- cbind(A,b)
}

# Create upper triangle matrix
# Note - doesn't handle rows with 0 pivots
triangle <- function(Aug){
    for (i in 2:nrow(Aug)){
        for(j in 1:(i - 1)){
            Aug[i,] <-  Aug[i,] - (Aug[j,] * (Aug[i, j]/Aug[j, j])) 
        }
    }
    Aug
}

# Backsolve upper triangle matrix, return vector x
backsolve <- function(Uc){
  x <- c(NA*3)
  x[3] <- Uc[3,4] / Uc[3,3]
  x[2] <- (Uc[2,4] - Uc[2,3] * x [3]) / Uc[2,2]
  x[1] <- (Uc[1,4] - Uc[1,2] * x[2] - Uc[1,3] * x[3]) / Uc[1,1]
  x <- round(x, 2)
  x
}
```

## Function to Solve

```{r}
#function to call all helpers, produce x that solves Ax = b for 3*3 matrix w/ no zero pivots
solve3by3nozeropivot <- function(A,b){
    Ab <- combine(A,b)
    Uc <- triangle(Ab)
    x <- backsolve(Uc)
    x
}
solve3by3nozeropivot(A,b)
```

## Double Check

```{r}
A <- matrix(c(2,4,-2,4,9,-3,-2,-3,7), 3, byrow=T)
b <- matrix(c(2,8,10))
solve3by3nozeropivot(A,b)
```