---
title: "DATA 605: Fundamentals of Computational Mathematics"
author: "Walt Wells, Spring 2017"
subtitle: "Assignment 2"
output:
  html_document:
    css: ../custom.css
    highlight: zenburn
    theme: lumen
---
# Problem 1

__1. Show that $A^T A \neq AA^T$ in general. (Proof and demonstration.)__

_Indirect Proof:_

* Assume $A^T A = A A^T$
* If A = $\begin{bmatrix} 2 & 2 \\3 & 4\end{bmatrix}$ then $A^T$ = $\begin{bmatrix} 2 & 3 \\2 & 4\end{bmatrix}$
* $A^T A = \begin{bmatrix} 2 & 3 \\2 & 4\end{bmatrix}\begin{bmatrix} 2 & 2 \\3 & 4\end{bmatrix} = \begin{bmatrix} 13 & 16 \\16 & 20\end{bmatrix}$
* but $A A^T = \begin{bmatrix} 2 & 2 \\3 & 4\end{bmatrix}\begin{bmatrix} 2 & 3 \\2 & 4\end{bmatrix} = \begin{bmatrix} 8 & 14 \\14 & 25\end{bmatrix}$
* $\therefore$ we conclude that the original proposition is false and in general $A^T A \neq AA^T$

__2. For a special type of square matrix A, we get $A^T A = AA^T$ . Under what conditions could this be true? (Hint: The Identity matrix I is an example of such a matrix).__

The commutative law is usually broken when attempting to multiply a matrix by its transposition, but exceptions are made for symmetric matrices.   We can obtain a symmetric matrix by multiplying any $A$ by $A^T$ or $A^T$ by any $A$ to get $S$.  Then since $S^T = S$ there is no difference in the resulting dot product.

```{r}
A <- matrix(rnorm(9), 3)
S <- t(A) %*% A
S %*% t(S)  
t(S) %*% S

```


# Problem 2

Matrix factorization is a very important problem. There are supercomputers built just to do matrix factorizations. Every second you are on an airplane, matrices are being factorized. Radars that track flights use a technique called Kalman filtering. At the heart of Kalman Filtering is a Matrix Factorization operation. Kalman Filters are solving linear systems of equations when they track your flight using radars.

Write an R function to factorize a square matrix A into LU or LDU, whichever you prefer. You don't have to worry about permuting rows of A and you can assume that A is less than 5x5, if you need to hard-code any variables in your code. 

## Testing Matrices

```{r}
#from week2mod1
A <- matrix(c(1,2,3,1,1,1,2,0,1),nrow=3)

E21 <- matrix(c(1,-2,0,0,1,0,0,0,1),nrow=3)
E31 <- matrix(c(1,0,-3,0,1,0,0,0,1),nrow=3)
E32 <- matrix(c(1,0,0,0,1,-2,0,0,1),nrow=3)

U <- E32 %*% E31 %*% E21 %*% A
L <- solve(E21) %*% solve(E31) %*% solve(E32)
```

## Helper Functions

```{r}
# get dimensions of given matrix
sizer <- function(matrix){
    i <- nrow(matrix)
    j <- ncol(matrix)
    dim <- c(i,j)
    dim
}

#set identity matrix given dim
identity <- function(dim){
    I <- diag(dim[1])
    I
}

#test to confirm LU = A
tester <- function(list, M){
    print("L:")
    print(list[[1]])
    print("U:")
    print(list[[2]])
    print("LU:")
    LU <- list[[1]] %*% list[[2]] 
    print(LU)
    print("A:")
    print(M)
}

```

## Function to Solve A = LU

```{r}
LowerUpper <- function(A, debug=FALSE){
    
    #get matrix dimensions
    dim <- sizer(A)
    
    ## error handling for non-square submissions
    if(dim[1]!=dim[2]) {
        return("This is not a square matrix")
    }
    
    ## initialize our L as an identity matrix to store successive E^-1
    ## initialize our U as a copy of A
    L <- identity(dim)
    U <- A
    
    ## solve U based on E, store E^-1 in L
    for (i in 2:dim[1]){
        for(j in 1:(i - 1)){
            # initialize a new E for each iteration
            E <-  identity(dim)
            # Find E based on state of U
            E[i, j] <- -(U[i,j]/U[j,j])
            
            # update U using E
            U <- E %*% U
            # update L based on E^-1
            L <- L %*% solve(E)
            
            ## debugger to check status of matrices at each pivot
            if (debug==TRUE){
                print(sprintf("E_%s%s:", i, j))
                print(E)
                print(sprintf("U_%s%s:", i, j))
                print(U)            
                print(sprintf("L_%s%s:", i, j))
                print(L)
            }
        }
    }
    
    return(list(L,U))
}
```

## Function Calls / Test

```{r}
# from strang "intro to linear algebra" textbook
b <- matrix(c(1,2,4,9), 2, byrow=T)
t1 <- LowerUpper(b)
tester(t1, b)

t2 <- LowerUpper(A)
tester(t2, A)

pascal <- matrix(c(1,1,1,1,
                   1,2,3,4,
                   1,3,6,10,
                   1,4,10,20), 4, byrow=T)
t3 <- LowerUpper(pascal)
tester(t3, pascal)

d <- matrix(seq(1:6), 3, byrow=T)
t6 <- LowerUpper(d)
t6
```